model_fit

### 2025-12-14 Draft2.Rmd
Paper update1

## Motivation
- Bank marketing dynamics (cite the paper)

@@

- Introduce dataset
    - Description of variables (under factors(
    - Description of target variable (EDA of several factors with target; plots) 
    - note imbalance 
    - Include wrangling of dataset (as.factor() and addition of variable for pdays)

Our dataset was a publicily available daaset, thus had less variables for privacy concerns of the bank customers...
The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts. A list of the 16 variable found below...
The response variable indicates whether a client subscribed to a term deposit which defines a binary classification task.

@@

@@

- Problem Statement
    - introduction of ML (as in ISLR)
        - Chapter 2: Estimating f(x)
            - How some models are more flexible, or more accurate. How some are less interpretable 
            - Assessing model accuracy
        - Chapter 4: We are dealing with classification problem

@@

## Methodology
For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets.

1. Logistic regression 
    * Describe fitting algorithm (from ISLR)
    * code form model_fir
    * short paragraph after each action
        * “we did this because…” or “we didn’t do this because…”
Logistic regression is used to model the probability of a binary outcome as a function of predictor variables. The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$

Formula: f(x) =

Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data.

Feature selection is performed using backward stepwise selection to identify a parsimonious subset of predictors that minimizes estimated test error as measured by AIC. Variables such as duration and previous marketing outcomes remain in the model due to their strong predictive power, whereas less informative variables are excluded to reduce variance. Categorical predictors are converted to factors, and a new binary variable previously_contacted is created to handle clients with no prior campaign contact. This preprocessing ensures that the data is suitable for statistical modeling and facilitates accurate interpretation of model coefficients.

Next, we check for potential non-linearity in continuous predictors from our predictors remaining after back-ward stepwise selection. We plotted deviance residuals against numeric features.
Based on residual diagnostics, we incorporated a polynomial term for balance to account for non-linear effects.

Regularization techniques such as Ridge (l2 penalty) and LASSO (l1 penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors. Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error. Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority.

Ridge and LASSO models yield similar results, indicating that additional shrinkage does not meaningfully improve classification of subscribers. We assume this is due to our modeling being majority indicator variables. This binary aspect of the indicator variable causes it's coefficient to be less affected by regularization techniques.
The ROC curves and AUC values of approximately 0.90 confirm that all models perform similarly in terms of discrimination between the two classes. These findings suggest that the primary limitation is bias from model flexibility rather than variance, and highlight the need for potentially more flexible classifiers to improve minority class prediction.

  2. KNN
  3. Boasted trees

>>>>

```{r, include=FALSE}
library(readr)
library(tidyverse)
library(boot)
library(glmnet)
library(caret)
library(pROC)
library(MASS)
library(ggplot2)
library(dplyr)

set.seed(445)

bank_full <- read_delim("data/bank-full_1.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

bank_full$y <- factor(bank_full$y, levels = c("no", "yes"))

bank_full <- bank_full %>%
  mutate(
    job = as.factor(job),
    marital = as.factor(marital),
    education = as.factor(education),
    default = as.factor(default),
    housing = as.factor(housing),
    loan = as.factor(loan),
    contact = as.factor(contact),
    month = as.factor(month),
    poutcome = as.factor(poutcome)) %>% 
  mutate(previously_contacted = as.factor(ifelse(pdays == -1, "no", "yes")))

## Train/Test split

train_index <- createDataPartition(bank_full$y, p = 0.8, list = FALSE, times = 1)
train_data <- bank_full[train_index, ]
test_data  <- bank_full[-train_index, ]
```

## Motivation
- **Bank marketing dynamics (cite the paper)**

@@

- **Introduce dataset**
    - Description of variables (under factors(
    - Description of target variable (EDA of several factors with target; plots) 
    - note imbalance 
    - Include wrangling of dataset (as.factor() and addition of variable for pdays)

Our dataset was a publicily available daaset, thus had less variables for privacy concerns of the bank customers.
The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns. These campaigns were done to encourage clients to subscribe to a long-term deposit product. Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts. A list of the 16 variable found below...

* age: Age in years 
* job: Occupation (Categorical)
* marital: Marital status (Categorical)
* education: Highest level of education (Categorical)
* default: has credit in default? (binary)
* balance: average yearly balance
* housing: has housing loan? (binary)
* loan: has personal loan? (binary)
* Contract: contact communication type (Categorical)
* day: last contact day (of the month)
* month: last contact month of year (month; 1=January)
* duration: last contact duration, in seconds
* campaign: number of contacts performed during this campaign and for this client
* pdays: number of days that passed by after the client was last contacted from a previous campaign (-1 means client was not previously contacted)
* previous: number of contacts performed before this campaign and for this client
* poutcome: outcome of the previous marketing campaign (Categorical)
* y: Indicator of whether the client subscribed a term deposit (Binary target)

The response variable indicates whether a client subscribed to a term deposit which defines a binary classification task.

@@

*[STREAMLINE Plots and EDA summary HERE]*

@@

- **Problem Statement**
    - introduction of ML (as in ISLR)
        - Chapter 2: Estimating f(x)
            - How some models are more flexible, or more accurate. How some are less interpretable 
            - Assessing model accuracy
        - Chapter 4: We are dealing with classification problem

@@

## Methodology
For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets. Our models do not apply time-series modeling, since our data does not include sequintial variables. Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series. This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.

1. **Logistic regression**

```{r, include=FALSE}
full_formula <- y ~ age + job + marital + education + default + balance + 
  housing + loan + contact + day + month + duration + campaign + 
  pdays + previous + poutcome + previously_contacted

glm.full <- glm(full_formula, data = train_data, family = "binomial")

# Perform backward selection
glm.step <- step(glm.full, direction = "backward", trace = 0)

selected_formula <- formula(glm.step)

## 
glm.baseline <- glm(selected_formula, data = train_data, family = "binomial")
summary(glm.baseline)

ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE, 
  summaryFunction = twoClassSummary 
)

set.seed(445) 
cv_fit <- train(
  selected_formula, 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl, 
  metric = "ROC" 
)

test_probabilities <- predict(glm.baseline, newdata = test_data, type = "response")
test_predictions <- factor(ifelse(test_probabilities > 0.5, "yes", "no"), levels = levels(test_data$y))

conf_matrix <- confusionMatrix(data = test_predictions, reference = test_data$y, positive = "yes" )

# Extract deviance residuals
residuals <- residuals(glm.baseline, type = "deviance")
train_data$residuals <- residuals

p1 <- ggplot(train_data, aes(x = balance, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Deviance Residuals vs. Balance",
       y = "Deviance Residuals") +
  theme_minimal()

glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
summary(glm.tran)
tran.coef <- summary(glm.tran)$coef

residuals <- residuals(glm.tran, type = "deviance")
train_data$residuals <- residuals

p2 <- ggplot(train_data, aes(x = balance, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Deviance Residuals vs. Balance",
       y = "Deviance Residuals") +
  theme_minimal()

## Regularization
X_full <- model.matrix(y ~ . - 1, data = bank_full)
Y_full <- as.numeric(bank_full$y) - 1

X_train <- X_full[train_index, ]
Y_train <- Y_full[train_index]

X_test <- X_full[-train_index, ]
Y_test <- Y_full[-train_index]

cv.ridge <- cv.glmnet(X_train, Y_train, 
                      alpha = 0, 
                      family = "binomial", 
                      type.measure = "class")
lambda.min.ridge <- cv.ridge$lambda.min

cv.lasso <- cv.glmnet(X_train, Y_train, 
                      alpha = 1, 
                      family = "binomial", 
                      type.measure = "class")
lambda.min.lasso <- cv.lasso$lambda.min

baseline_prob <- predict(glm.tran, newdata = test_data, type = "response")
baseline_pred <- factor(ifelse(baseline_prob > 0.5, "yes", "no"), levels = c("no", "yes"))

ridge_prob <- predict(cv.ridge, s = lambda.min.ridge, newx = X_test, type = "response")
ridge_pred <- factor(ifelse(ridge_prob > 0.5, "yes", "no"), levels = c("no", "yes"))

lasso_prob <- predict(cv.lasso, s = lambda.min.lasso, newx = X_test, type = "response")
lasso_pred <- factor(ifelse(lasso_prob > 0.5, "yes", "no"), levels = c("no", "yes"))

cm_baseline <- confusionMatrix(baseline_pred, test_data$y, positive = "yes")
cm_ridge <- confusionMatrix(ridge_pred, test_data$y, positive = "yes")
cm_lasso <- confusionMatrix(lasso_pred, test_data$y, positive = "yes")

metrics <- data.frame(
  Model = c("Baseline Logit", "Ridge Logit", "LASSO Logit"),
  Accuracy = c(cm_baseline$overall['Accuracy'], cm_ridge$overall['Accuracy'], cm_lasso$overall['Accuracy']),
  Sensitivity = c(cm_baseline$byClass['Sensitivity'], cm_ridge$byClass['Sensitivity'], cm_lasso$byClass['Sensitivity']),
  Specificity = c(cm_baseline$byClass['Specificity'], cm_ridge$byClass['Specificity'], cm_lasso$byClass['Specificity'])
)

roc_baseline <- roc(test_data$y, baseline_prob)
roc_ridge <- roc(test_data$y, as.numeric(ridge_prob))
roc_lasso <- roc(test_data$y, as.numeric(lasso_prob))

roc_data <- data.frame(
  fpr = c(1 - roc_baseline$specificities, 1 - roc_ridge$specificities, 1 - roc_lasso$specificities),
  tpr = c(roc_baseline$sensitivities, roc_ridge$sensitivities, roc_lasso$sensitivities),
  Model = factor(c(rep("Baseline Logit", length(roc_baseline$specificities)),
                   rep("Ridge Logit", length(roc_ridge$specificities)),
                   rep("LASSO Logit", length(roc_lasso$specificities))),
                 levels = c("Baseline Logit", "Ridge Logit", "LASSO Logit"))
)

```

Logistic regression is used to model the probability of a binary outcome as a function of predictor variables. The model estimates the probability that a client subscribes to a term deposit, $P(Y=1 \mid X)$. This model fits with assumptions of independent observations and normality.

Formula: $$f(x) = P(Y=1 \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \dots + \beta_k x_k)}}$$  
Coefficients are estimated using the maximum likelihood method, which finds the parameter values that maximize the likelihood of observing the training data (ISLR).

Feature selection is performed using backward stepwise selection to identify a parsimonious subset of predictors that minimizes estimated test error as measured by AIC. Variables such as duration and previous marketing outcomes remain in the model due to their strong predictive power, whereas less informative variables are excluded to reduce variance. Categorical predictors are converted to factors, and a new binary variable previously_contacted is created to handle clients with no prior campaign contact. This preprocessing ensures that the data is suitable for statistical modeling and facilitates accurate interpretation of model coefficients. We resulted 12 features:

```{r}
selected_formula
```

Cross validation is used to select optimal tuning parameters for these regularized models and to estimate the expected test error. Confusion matrices, sensitivity, specificity, and ROC/AUC metrics are computed to assess predictive performance, especially given the class imbalance where non-subscribers are the majority. This model performed will with mean 5-fold CV test error metrics:

```{r, echo=FALSE}
print(conf_matrix)
```

Next, we check for potential non-linearity in continuous predictors from our predictors remaining after back-ward stepwise selection. We plotted deviance residuals against numeric features.
Based on residual diagnostics, we incorporated a polynomial term (using the poly function: *poly(balance, 4)*) for balance to account for non-linear effects. The Balance squared and power 4 term produced significant coefficients (Last 4 on the list of coefficients).

```{r, out.width="45%", echo=FALSE}
p1
p2
```

```{r}
tran.coef
```

Although this polynomial transformation didn't provide desired metrics in expense of the nonlinear effect. From the deviance residual plots we can observe there weren't significant deviance from a normal residual line. We will not include this model in our comparison.

Regularization techniques such as Ridge (l2 penalty) and LASSO (l1 penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors. LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$). A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6). The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`. Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.

```{r, out.width="45%", echo=FALSE}
plot(cv.ridge)
plot(cv.lasso)
```

Here is a comparison of the models fitted thus far:

```{r}
metrics
```

Ridge and LASSO models yield similar results, indicating that additional shrinkage does not meaningfully improve classification of subscribers. We assume this is due to our modeling being majority indicator variables. This binary aspect of the indicator variable causes it's coefficient to be less affected by regularization techniques.

```{r, out.width="45%", echo=FALSE}
ggplot(roc_data, aes(x = fpr, y = tpr, color = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(linetype = "dashed", color = "grey") +
  labs(
    title = "ROC Curve Comparison for Logistic Models",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()
```

All three models exhibit a high overall Accuracy (around 0.89), which is expected given the significant class imbalance (Prevalence $\approx 0.11$). However, the critical metric, Sensitivity (the True Positive Rate for the minority class, 'yes'), is very low (around 0.17). This result, consistent across all three models, indicates that they struggle to correctly identify subscribers, leading to many False Negatives (FN). Conversely, the high Specificity (True Negative Rate $\approx 0.98$) means the models are excellent at identifying non-subscribers. The regularization methods (Ridge and LASSO) failed to provide any significant improvement in these metrics, suggesting that the primary limitation is one of high bias (due to insufficient flexibility) rather than high variance.
\

  2. **KNN** 
      - Describe fitting algorithm (from ISLR)
      - sample code and plots form model_fit file
      - short paragraph after each action (“we did this because…” or “we didn’t do this because…”) \
  3. **Boasted trees** 
      - Describe fitting algorithm (from ISLR)
      - sample code and plots form model_fit file
      - short paragraph after each action (“we did this because…” or “we didn’t do this because…”) \

### 2025-12-14 model_fit.Rmd
Paper update1

@@

```{r}
glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
summary(glm.tran)

@@

ggplot(train_data, aes(x = balance, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Deviance Residuals vs. Balance",
       y = "Deviance Residuals") +
  theme_minimal()
```

@@

The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, r round(lambda.min.lasso, 5). Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.

>>>>

```{r}
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE, 
  summaryFunction = twoClassSummary 
)

set.seed(445)
cv_fit <- train(
  selected_formula, 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl, 
  metric = "ROC" 
)

print(cv_fit)

test_probabilities <- predict(
  glm.baseline, 
  newdata = test_data, 
  type = "response"
)

test_predictions <- factor(
  ifelse(test_probabilities > 0.5, "yes", "no"),
  levels = levels(test_data$y)
)

conf_matrix <- confusionMatrix(
  data = test_predictions, 
  reference = test_data$y, 
  positive = "yes" 
)

print(conf_matrix)

@@

```{r}
glm.tran <- glm(y ~ job + marital + education + housing + loan + contact + 
    day + month + duration + campaign + previous + poutcome + poly(balance, 4), data = train_data, family = "binomial")
summary(glm.tran)
tran.coef <- summary(glm.tran)$coef

@@

p2 <- ggplot(train_data, aes(x = balance, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Deviance Residuals vs. Balance",
       y = "Deviance Residuals") +
  theme_minimal()
```

@@

The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`. Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.
